A method for regularization of evolutionary polynomial regressions

### 2013-12-24 texts/EPRR_FILES-ASC/eprr.tex
Setup to submit to “Applied Soft Computing”.

%\documentclass[review,preprint,authoryear,12pt]{elsarticle}
\documentclass[preprint,authoryear,12pt]{elsarticle}

%
%	LINE NUMBERING FOR REVISION
%
%\usepackage{lineno}
%

%
%	FILE ENCODING
%
\usepackage[utf8]{inputenc}
%

%
%	PAGE GEOMETRY
%
%\usepackage{geometry}
%\geometry{a4paper}
%

%
%	AMS STUFF
%
\usepackage{amsfonts}
%\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
%

%
%	ACRONYMS
%
\usepackage{acronym}
%
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square}
%\acrodef{EPR}[\textsc{EPR}]{Genetic Algorithms for Polynomials}
%\acrodef{EPRR}[\textsc{EPR.reg}]{Genetic Algorithms for Polynomials with Regularization}

\acrodef{EPR}[EPR]{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPR$^2$]{Evolutionary Polynomial Regression with Regularization}

\acrodef{EPR}{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPRR]{Evolutionary Polynomial Regression with Regularization}

\acrodef{SVM}[SVM]{Support Vector Machines}
%

%
%	FIGURE AND STUFF ENVIRONMENT
%
\usepackage{graphicx}
%

%
%	ALGORITHMS PACKAGES
%
\usepackage{algpseudocode}
\usepackage{algorithm}
%

%
%	HYPERREFERENCES
%
\usepackage{hyperref}
%

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

%
%	DOCUMENT COMMANDS
%
\newcommand{\sample}[1]{\ensuremath{^{\left(#1\right)}}}
\newcommand{\at}[1]{\ensuremath{\!\left(#1\right)}}
\newcommand{\note}[1]{\textbf{\small{[Note: {#1}]}}}
\newcommand{\nth}{\ensuremath{^{th}}}
%

%
%
%%% BEGIN DOCUMENT
%
%
\begin{document}
\begin{frontmatter}

\title{A Method for Regularization of Evolutionary Polynomial Regressions}

\author[ue,labmag]{Francisco ~Coelho\corref{fc}}
\ead{fc@di.uevora.pt}
\cortext[fc]{Corresponding author}

\author[fcul,labmag]{João ~Pedro ~Neto}
\ead{jpn@di.fc.ul.pt}

\address[ue]{Dept. Informática, Universidade de Évora, Rua Romão Ramalho 58, 7000-671 Évora}
\address[fcul]{Dept. Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande 1749-016 Lisboa}
\address[labmag]{Laboratory of Agent Modelling (LabMAg)}
%
%\date{[{\sc draft} \today]}
%
\begin{abstract}
While many applications require models that have no acceptable linear approximation, the simpler nonlinear models are defined by polynomials. The use of genetic algorithms to find polynomial models from data is known as \ac{EPR}.
%
This paper introduces \ac{EPRR}, an algorithm that extends the \ac{EPR} method and describes a set of experiences on common datasets that compare both flavors of \ac{EPR} and other methods including Linear Regression, Regression Trees and Support Vector Regression.

The empiric conclusion of those experiments is that \ac{EPRR} is able to achieve better fitting than other non-ensemble methods and it has shorter computation time than plain \ac{EPR}.
\end{abstract}
%
\begin{keyword}
evolutionary polynomial regression \sep regularization \sep feature extraction \sep dimensionality reduction 
\end{keyword}
\end{frontmatter}
%
%	Line numbers for revision
%
%
%\linenumbers
%
\section{Introduction}

With notable exceptions (\emph{e.g.} neural networks) machine learning regression techniques produce linear models. The linearity assumption has many advantages including reduced computational complexity and strong the\-o\-re\-ti\-cal framework. However nonlinearity is unavoidable in many application scenarios, specially those with phase transitions or feedback loops, so common in engineering, ecology, cybernetics and other areas. The kernel trick in \ac{SVM} (\cite{scholkopf1997kernel, liang2012eigen, Bao:2013aa}) alleviates this problem by allowing special non-linear transformations of the feature-space. The condition such transformations must meet is known as the \emph{kernel trick}, $k\at{x,x'} = \left< \varphi\at{x}, \varphi\at{x'} \right>$, where $\varphi$ is the feature-space transformation and $\left<\cdot,\cdot\right>$ denotes inner product. The ``trick'' consists on computing the kernel $k\at{x,x'}$ while avoiding the computation of the inner product and the transformations $\varphi\at{x}, \varphi\at{x'}$. A special case of polynomial transformation, the \emph{polynomial kernel}, $k\at{x,x'} = \left<x,x'\right>^d$ is commonly used in regression and classification tasks with \acp{SVM}. However general polynomial transformations do not verify the kernel trick.

@@

The idea of using \acp{GA} to find a polynomial regression is not new \citep{Maertens:2006aa, Yu:2008aa, Wu:2009aa} but still generates original research \citep{Hofwing:2011aa,Cetisli:2011aa}. The modern formulation of the use of \ac{GA} to find polynomial models is known as \acf{EPR} and systematization can be traced back to the work of Davidson, Savic and Walters (\cite{Davidson:2003aa}). Further developments include multi-objective optimizations (\cite{Giustolisi:2009aa}).

@@

%A method that produces \emph{adequate} models from observed complex data has many uses. For example by a scientist to better understand the source of the data or by an autonomous agents adapting to the environment.

% It turns out that, discarding the computational cost of training, the regression method presented here, \acf{EPRR}, proves to be quite effective. Indeed, according to the results of common  methods on several real-world datasets it is only systematically out-performed by random forests, an \emph{ensemble} method.

@@

\section{Genetic Algorithms for Polynomials}

This section starts with a brief introduction and outline of the evolutionary polynomial regression algorithm, \ac{EPR}, and proceeds into core details as the encoding used to represent individual polynomial instances in the \ac{GA} populations and the regularization of the cost function.

@@

The exponents alone can be organized into a matrix $\left[\alpha_{ij}\right]$ that defines the monomial structure of the polynomial. For the example above the matrix representation of the monomials is
$$
\left[ 
\begin{array}{ccc}
1 & 0 & 0 \\
0 & 1 & 1 \\
2 & 0 & 1
\end{array}
\right] \sim 
\left[ 
\begin{array}{ccc}
x_1 \\
x_2 x_3 \\
x_1^2 x_3
\end{array}
\right]
$$
where each row defines a monomial and each column represents a variable. Changing the order of the rows doesn't change the polynomial whereas changing the order of the columns corresponds to changing the respective variables.
%Notice that if we interpret matrix $A$ as a single column of monomials and set the coefficients $T = \left[ \begin{array}{ccc} 2 & 1 & \frac{1}{2} \end{array} \right ]$ then
%$$
%p\left( x_1, x_2, x_3 \right) = TA = \left[ \begin{array}{ccc} 2 & 1 & \frac{1}{2} \end{array} \right ]\left[ 
%\begin{array}{ccc}
%x_1 \\
%x_2 x_3 \\
%x_1^2 x_3
%\end{array}
%\right].
%$$

@@

%
\subsection{Polynomial Encoding}\label{subs:polynomial.encoding}
%
%\begin{quotation}
%\note{Next paragraph moved here from the introduction.}
%Another important change is in the coding mechanism which introduces two new features: (a) inclusion of the linear terms in the polynomial expression, and (b) the activation/deactivation of monomials during evolution to allow the increase/decrease of polynomial expressions.
%\end{quotation}
%
The specific encoding (representation) of a set of monomials is an important aspect in the implementation of \ac{EPR}. The choice described below permits active and inactive monomials for regression purposes. The active (or inactive) state of a monomial might change through mutation or crossover. This simple mechanism enhances variation in the complexity of polynomial expressions by evolutionary operations.

@@

There is one final remark concerning this encoding method. As it is, the activation segment can become all zeros, representing the empty set of monomials. This situation can be avoided with a simple hack: Given an encoding, the first monomial is always considered active, thus restricting the syntactic form of encodings to binary strings starting with $1$. In practice, this means that the implementation of the encoding can omit the first bit.
% If the coding consists of entirely zeros, by convention, it describes polynomial $x_1$. This has to do with the execution of the polynomial regression that would fail if we interpret it as the zero polynomial. Anyway, for progressive larger  binary descriptions, the chances of getting this zero description decrease exponentially, so it does not impact in any meaningful way in the algorithm's performance.

@@

% best polynomial, rsme 0.65
% $$y = 0.52 s_1^2 x_5 - 0.13 x_2 x_8 -0.75 x_1^2 x_6 + 0.78 x_8 + 0.18 $$

@@

% best polynomial, \ac{rmse} 0.32
%$$ y = 0.4 x_6 + 0.02 x_5^3 x_6 - 0.85 x_4 + 0.23 x_4^2 - 0.26 $$

@@

% best polynomial, \ac{rmse} 0.74
% $$ y = 0.24 x_5 - 0.23 x_5 x_6 x_7 - 0.19 x_4 x_7 - 0.5 x_3 $$

@@

%Since the entire learning process takes some time, in the current \texttt{R} implementation, placing a limit between 50 to 100 generations already achieves good results, relative to higher iteration values. Figure~\ref{Abalone_fitnessProgress} shows a typical error evolution for the dataset \textsc{Abalone} given a regularized version ($\lambda = 0.975$) and a non-regularized ($\lambda = 1.0$).\note{Why not $\lambda=0.8$?}

@@

\section*{Bibliography}

\bibliographystyle{model2-names}
\bibliography{fullbib2}

>>>>

%% Use the option review to obtain double line spacing
\documentclass[times,review,preprint]{elsarticle}

%% Use the options `twocolumn,final' to obtain the final layout
%\documentclass[times,twocolumn,final,authoryear]{elsarticle}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
%
%	ALGORITHMS PACKAGES
%
\usepackage{algpseudocode}
\usepackage{algorithm}
%
%	HYPERREFERENCES
%
\usepackage{url}
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%	DOCUMENT COMMANDS
%
\newcommand{\sample}[1]{\ensuremath{^{\left(#1\right)}}}
\newcommand{\at}[1]{\ensuremath{\!\left(#1\right)}}
\newcommand{\note}[1]{\textbf{\small{[Note: {#1}]}}}
\newcommand{\nth}{\ensuremath{^{th}}}
%
\usepackage{lineno}
%
%	ACRONYMS
%
\usepackage{acronym}
%
%
%
\def\TM{$^{\rm TM}$}

\journal{Applied Soft Computing}

\begin{document}

%
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square}
\acrodef{EPR}[EPR]{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPR$^2$]{Evolutionary Polynomial Regression with Regularization}
\acrodef{EPR}{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPRR]{Evolutionary Polynomial Regression with Regularization}
\acrodef{SVM}[SVM]{Support Vector Machines}

\setcounter{page}{0}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{A method for regularization of evolutionary polynomial regressions}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[1]{Francisco Coelho\corref{cor1}} 
\cortext[cor1]{Corresponding author: 
  Tel.: +351-919-006-379}
\ead{fc@di.uevora.pt}
\author[2]{Jo\~{a}o Pedro Neto}

\address[1]{Departamento de Informática, Universidade de Évora, Rua Romão Ramalho 58, 7000-671 Évora, Portugal}
\address[2]{Departamento de Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande, 1749-016 Lisboa, Portugal}

\begin{abstract}
While many applications require models that have no acceptable linear approximation, the simpler nonlinear models are defined by polynomials. The use of genetic algorithms to find polynomial models from data is known as Evolutionary Polynomial Regression.
%
This paper introduces Evolutionary Polynomial Regression with Regularization, an algorithm that extends the EPR method and describes a set of experiences on common datasets that compare both flavors of EPR and other methods including Linear Regression, Regression Trees and Support Vector Regression.

The empiric conclusion of those experiments is that EPR with regularization is able to achieve better fitting than other non-ensemble methods and it has shorter computation time than plain EPR.
\end{abstract}

\begin{keyword}
evolutionary polynomial regression \sep regularization \sep feature extraction \end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
With notable exceptions (\emph{e.g.} neural networks) machine learning regression techniques produce linear models. The linearity assumption has many advantages including reduced computational complexity and strong the\-o\-re\-ti\-cal framework. However nonlinearity is unavoidable in many application scenarios, specially those with phase transitions or feedback loops, so common in engineering, ecology, cybernetics and other areas. The kernel trick in \ac{SVM} (\cite{scholkopf1997kernel, liang2012eigen, Bao:2013aa}) alleviates this problem by allowing special non-linear transformations of the feature-space. The condition such transformations must meet is known as the \emph{kernel trick}, $k\at{x,x'} = \left< \varphi\at{x}, \varphi\at{x'} \right>$, where $\varphi$ is the feature-space transformation and $\left<\cdot,\cdot\right>$ denotes inner product. The ``trick'' consists on computing the kernel $k\at{x,x'}$ while avoiding the computation of the inner product and the transformations $\varphi\at{x}, \varphi\at{x'}$. A special case of polynomial transformation, the \emph{polynomial kernel}, $k\at{x,x'} = \left<x,x'\right>^d$ is commonly used in regression and classification tasks with \acp{SVM}. However general polynomial transformations do not verify the kernel trick.

@@

The idea of using \acp{GA} to find a polynomial regression is not new \citep{Maertens:2006aa, Yu:2008aa, Wu:2009aa} but still generates original research \citep{Hofwing:2011aa,Cetisli:2011aa}. The modern formulation of the use of \ac{GA} to find polynomial models is known as \acf{EPR} and systematization can be traced back to the work of Davidson, Savic and Walters \cite{Davidson:2003aa}. Further developments include multi-objective optimizations \cite{Giustolisi:2009aa}.

@@

@@

\section{Genetic Algorithms for Polynomials}
This section starts with a brief introduction and outline of the evolutionary polynomial regression algorithm, \ac{EPR}, and proceeds into core details as the encoding used to represent individual polynomial instances in the \ac{GA} populations and the regularization of the cost function.

@@

The exponents alone can be organized into a matrix $\left[\alpha_{ij}\right]$ that defines the monomial structure of the polynomial. For the example above the matrix representation of the monomials is
$$
\left[ 
\begin{array}{ccc}
1 & 0 & 0 \\
0 & 1 & 1 \\
2 & 0 & 1
\end{array}
\right] \sim 
\left[ 
\begin{array}{ccc}
x_1 \\
x_2 x_3 \\
x_1^2 x_3
\end{array}
\right]
$$
where each row defines a monomial and each column represents a variable. Changing the order of the rows doesn't change the polynomial whereas changing the order of the columns corresponds to changing the respective variables.

@@

%
\subsection{Polynomial Encoding}\label{subs:polynomial.encoding}

The specific encoding (representation) of a set of monomials is an important aspect in the implementation of \ac{EPR}. The choice described below permits active and inactive monomials for regression purposes. The active (or inactive) state of a monomial might change through mutation or crossover. This simple mechanism enhances variation in the complexity of polynomial expressions by evolutionary operations.

@@

There is one final remark concerning this encoding method. As it is, the activation segment can become all zeros, representing the empty set of monomials. This situation can be avoided with a simple hack: Given an encoding, the first monomial is always considered active, thus restricting the syntactic form of encodings to binary strings starting with $1$. In practice, this means that the implementation of the encoding can omit the first bit.

@@

@@

@@

@@

@@

\bibliographystyle{elsarticle-num}
\bibliography{fullbib2}
